---
title: "SDS 291 Final Project: Topic and Data Selection"
author: "Miya Dang, Mia Tran, Alua Birgebayeva"
date: "Monday, April 28, 2025"
format: 
  pdf:
    keep-tex: true
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
    geometry: 
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
editor_options: 
  chunk_output_type: inline
---

# Data

The dataset we are using for the final project is the Heart Failure Clinical Records Dataset, originally sourced from the UCI Machine Learning Repository (<https://archive.ics.uci.edu/dataset/519/heart+failure+clinical+records>). It contains the medical records of 299 patients who experienced heart failure, collected during their follow-up period. Each observation is a patient profile, with 13 variables measuring their clinical records such as blood test results, whether or not they have diabetes or hypertension, as well as their age and sex.

# Research Question/Purpose

Our project aims to answer the following research question: **Which clinical characteristics are most predictive of death among patients with heart failure?**

We will identify important predictors that may offer insights into patient outcomes after heart failure. Then, using multiple logistic regression, we will model the probability of death during the follow-up period based on clinical data.

# Response Variable

Our response variable is DEATH_EVENT, a binary categorical variable with 2 levels:

-   0 = Patient survived during the follow-up period (considered "failure" when modelling).

-   1 = Patient died during the follow-up period (considered "success" when modelling).

# Explanatory Variables

We plan to include all 12 variables other than DEATH_EVENT as explanatory variables to include in the richest possible model .

| Variable                     | Type        | Description                                                    | Units/Levels                         |
|------------------|------------------|------------------|------------------|
| **age**                      | Numeric     | Age of the patient                                             | Years (40--95)                       |
| **anaemia**                  | Categorical | Presence of anemia (decrease of red blood cells or hemoglobin) | 0 = No, 1 = Yes                      |
| **creatinine_phosphokinase** | Numeric     | Level of CPK enzyme in blood (indicator of muscle damage)      | mcg/L (23--7861)                     |
| **diabetes**                 | Categorical | Presence of diabetes                                           | 0 = No, 1 = Yes                      |
| **ejection_fraction**        | Numeric     | Percentage of blood leaving the heart at each contraction      | \% (14--80)                          |
| **high_blood_pressure**      | Categorical | Presence of hypertension (high blood pressure)                 | 0 = No, 1 = Yes                      |
| **platelets**                | Numeric     | Amount of platelets in the blood                               | kiloplatelets/mL (\~25,000--850,000) |
| **serum_creatinine**         | Numeric     | Level of serum creatinine in blood (kidney function indicator) | mg/dL (0.5--9.4)                     |
| **serum_sodium**             | Numeric     | Level of serum sodium in blood                                 | mEq/L (113--148)                     |
| **sex**                      | Categorical | Gender of patient                                              | 0 = Woman, 1 = Man                   |
| **smoking**                  | Categorical | Smoking status of patient                                      | 0 = No, 1 = Yes                      |
| **time**                     | Numeric     | Length of follow-up period                                     | Days (4--285)                        |

# Exploratory Visualizations

```{r, echo = FALSE}
library(ggplot2)
heart_data <- read.csv("heart_failure_clinical_records_dataset.csv")
```

```{r, echo = FALSE}
# Boxplot for Ejection Fraction
ggplot(heart_data, aes(x = factor(DEATH_EVENT), y = ejection_fraction)) +
  geom_boxplot() +
  labs(x = "Death Event (0 = Alive, 1 = Dead)", y = "Ejection Fraction (%)",
       title = "Ejection Fraction by Death Event")
```

The first boxplot shows that ejection fraction was generally lower among patients who died compared to those who survived, suggesting reduced heart pumping function as a key predictor of mortality.

```{r, echo = FALSE}
# Boxplot for Serum Creatinine
ggplot(heart_data, aes(x = factor(DEATH_EVENT), y = serum_creatinine)) +
  geom_boxplot() +
  labs(x = "Death Event (0 = Alive, 1 = Dead)", y = "Serum Creatinine (mg/dL)",
       title = "Serum Creatinine by Death Event")
```

The second boxplot shows that serum creatinine levels were higher among patients who died, indicating that impaired kidney function is associated with worse outcomes in heart failure patients.
